---
title: "Intro to Reshaping with Separate/Unite and Pivot"
subtitle: "STT2860 Intro to Data Management and Vis S22"
author: "Author: Lauren Rusnak"
date: "`r format(Sys.time(), '%A, %B %d, %Y @ %I:%M %p')`"
output: 
  html_document: 
    theme: cosmo
    highlight: textmate
---

```{r globaloptions, include = FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  comment = NA
)
```
<hr>

This worksheet is meant to be used with:

* [RDS CH12 Tidy data](https://r4ds.had.co.nz/tidy-data.html)
* [dplyr: separate and unite](https://youtu.be/ztcsf7G4S78) (video)
* [How to reshape data with tidyr’s new pivot functions](https://youtu.be/gnUgSkKEW5c) (video)
* [Function reference for tidyr](https://tidyr.tidyverse.org/reference/index.html)

## Packages Used

```{r}
library(tidyverse)
library(skimr) # for skim function
```

## Datasets Used

Example datasets used in **RDS CH12** from the **tidyr** package.

```{r}
table1
table2
table3
table4a
table4b
```

Example dataset from Dr. Thomley's Class Data Files (linked from AsULearn).

```{r}
powerball <- read_csv("https://raw.githubusercontent.com/STAT-JET-ASU/Datasets/master/Instructor/powerballnumbers.csv")
head(powerball)
```

Example dataset originally from the Kaggle repo [Netflix Movies and TV Shows](https://www.kaggle.com/shivamb/netflix-shows).

```{r}
netflix <- read_csv("https://raw.githubusercontent.com/STAT-JET-ASU/Datasets/master/Kaggle/netflix_2021_10_01.csv")
head(netflix)
```

## Separating and Uniting Columns

### `separate()`

Separating the character `drawdata` variable into 3 character vectors for `month`, `day`, and `year`. 
- New variable if named the same as an og column will overwrite that og column.
- `convert` = TRUE means that the character value will be converted to the highest possible type the data can be, in this case the character is converted to integer


```{r}
powerball2 <- powerball %>%
  separate(
    col = drawdate, # which column to separate
    into = c("month", "day", "year"), # names of new columns, must be in quotes
    sep = "/", # how numbers/data is separated in the og column
    remove = TRUE, # Default is shown. to keep original column, use FALSE
    convert = FALSE, # Default so can erase
    extra = "warn", # Default so can erase
    fill = "warn" # Default so can erase
)
head(powerball2)
```

### `unite()`

Ex.  dataset has separate day, month and year, but we want one variable

```{r}
powerball3 <- powerball2 %>%
  unite(
    col = "datedraw", 
    month:year, # can use colon if columns are in order
    sep = "-"
  )
head(powerball3)
```

```{r}
powerball_united2 <- powerball2 %>%
  unite(
    col = "datedraw", 
    c(year, month, day), # can use colon if columns are in order
    sep = "-"
  )
head(powerball_united2)
```

## Pivoting the Data "Rectangle"

### `pivot_longer()`
Using `table4a` I want to make `year` a variable, I can work with in R. Rn the years are variable names. My final dataset should have 3 variable `country`, ` year`, and the values from `1999` and `2000` columns.

Afghanistan 1999 745
Afghanistan 2000 2666
etc...

```{r}
table4a

```

Now `year` will be a piece of info (variable), not variable names.

The old (outdated) function for this was `gather()`. It still works, but has been replaced by `pivot_longer`.

```{r}
table4a_long <- table4a %>%
  pivot_longer(
    cols = `1999` : `2000`, # what columns are getting stacked up
    names_to = "year",      # where will the column names go to
    values_to = "cases"     # what is the new name of the stacked data
  )
table4a_long
```

```{r}
table4b
```

```{r}
table4b_long <- table4b %>%
  pivot_longer(
    cols = `1999` : `2000`, # what columns are getting stacked up
    names_to = "year",      # where will the column names go to
    values_to = "population"     # what is the new name of the stacked data
  )

table4b_long
```


Now we join the two tables!

- `inner_join()`
- `left_join()`
- `right_join()`
- `full_join()`
- `semi_join()`
- `anti_join()`




```{r}
full_join(x = table4a_long, 
          y = table4b_long,
          by = c("country", "year")
          )
```


### `pivot_wider()`

```{r}
table2
```
 
Create a new column of `rate = cases / population` using `mutate()`

```{r}
table2_wider <- table2 %>%
  pivot_wider(
    names_from = type,    # my new columns will be named cases and population
    values_from = count,  # what colum values are going to fill my new columns
  ) %>%
  mutate(rate = cases / population * 10000)

table2_wider
```


## Netflix Example

### `separate()` movie categories

Let's first try four categories...

We'll look at a few rwos with `head()` and us `skim()` to find `NA` values. Keep the separated column for nowso we an verify our code.

```{r}
netflix_4cat <- netflix %>%
  separate(
    col = listed_in,
    into = c("cat1", "cat2", "cat3", "cat4"),
    sep = ",",
    remove = FALSE
  )
```

```{r}
head(netflix_4cat, n = 10)

netflix_4cat %>% 
  select(cat1:cat4) %>% 
  skim()
```



We can reduce the number of categories...
since it seems as if no show has a cat4

```{r}
netflix_3cat <- netflix %>%
  separate(
    col = listed_in,
    into = c("cat1", "cat2", "cat3"),
    sep = ",",
    remove = FALSE
  )
```

```{r}
head(netflix_3cat, n = 10)

netflix_3cat %>% 
  select(cat1:cat3) %>% 
  skim()
```


Use `pivot_longer()` to put all the categories in the same column.

```{r}
netflix_3cat_long <- netflix_3cat %>%
  pivot_longer(
    cols = c("cat1", "cat2", "cat3"),
    names_to = "cat_order",
    values_to = "cat_name"
  )

netflix_3cat_long %>%
  count(cat_name)
```

Remove any rows that have `NA` values in the new column you created.

Filter to get rid of rows that have NA as a `cat_name`

```{r}
netflix_3cat_long
```

<hr>
```{r}
sessionInfo()
```