---
title: "Intro to Binding and Joining in R"
subtitle: "STT2860 Intro to Data Management and Vis S22"
author: "Author: YOUR NAME HERE"
date: "`r format(Sys.time(), '%A, %B %d, %Y @ %I:%M %p')`"
output: 
  html_document: 
    theme: cosmo
    highlight: textmate
---

```{r globaloptions, include = FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  comment = NA
)
```
<hr>

This worksheet is meant to be used with:

* [RDS CH13 Relational data](https://r4ds.had.co.nz/relational-data.html)
* [Mutating joins](https://dplyr.tidyverse.org/reference/mutate-joins.html)
* [Filtering joins](https://dplyr.tidyverse.org/reference/filter-joins.html)
* [Dr. Thomley's Instructor Datasets](https://github.com/STAT-JET-ASU/Datasets/tree/master/Instructor)

## Packages Used

```{r packagesused}
library(tidyverse)
library(skimr)
```

## Datasets Used

### Myers-Briggs Personality Datasets

Personality types determined using the [Jung Typology Test](https://www.humanmetrics.com/personality).

#### STT1810 Classes

```{r readclass1}
class1 <- read_csv("https://raw.githubusercontent.com/STAT-JET-ASU/Datasets/master/Instructor/personalitySTT1810.csv")
glimpse(class1)
```

#### STT3850 Classes

```{r readclass2}
class2 <- read_csv("https://raw.githubusercontent.com/STAT-JET-ASU/Datasets/master/Instructor/personalitySTT3850.csv")
glimpse(class2)
```

### Census Population Data

Census data taken from tables found on Wikipedia.

#### 1970 and 1980

```{r readcensus1}
census1 <- read_csv("https://raw.githubusercontent.com/STAT-JET-ASU/Datasets/master/Instructor/census19701980.csv")
glimpse(census1)
```

#### 1990 and 2000

```{r readcensus2}
census2 <- read_csv("https://raw.githubusercontent.com/STAT-JET-ASU/Datasets/master/Instructor/census19902000.csv")
glimpse(census2)
```

#### 2010 and 2020

```{r readcensus3}
census3 <- read_csv("https://raw.githubusercontent.com/STAT-JET-ASU/Datasets/master/Instructor/census20102020.csv")
glimpse(census3)
```

### Electoral Votes

Data from [US Electoral College Votes Per State 1788-2020 | Kaggle](https://www.kaggle.com/datasets/daithibhard/us-electoral-college-votes-per-state-17882020)

```{r readelectoralvotes}
votes <- read_csv("https://raw.githubusercontent.com/STAT-JET-ASU/Datasets/master/Instructor/electoralvotes.csv")
glimpse(votes)
```

## Examples of Combining

### `bind_rows()`

#### Simple row binding

Want to row bind two classes. They have the same variables, and in each dataset one row represents one student.

The bind_rows() function matches on names. Are all names the same? If not, use rename `rename()` or similar operation.

```{r}
names(class1) == names(class2)
```

```{r}
classes <- bind_rows(class1, class2)

glimpse(classes)
```

```{r}
class1 %>%
  bind_rows(class2) # can also use with piping
```


#### Binding with an ID #1

If I do a simple binding, I don't know which data came from which class.
Us `.id` to get an id column so we know which class the data came from.

```{r}
classesID <- bind_rows(class1, class2, .id = "ClassID") # column that numbers the data sets called ClassID
glimpse(classesID)


```

#### Binding with an ID #2

If I want to specify my ID values, need to enter the datasets I want to bind as a named list. The names become the ID values

```{r}
classesID2 <- bind_rows(
  list(
    STT1810 = class1,
    STT3860 = class2
  ),
  .id = "ClassID")

glimpse(classesID2)
```

### `bind_cols()`

#### Simple column binding

What if I don't make sure my columns are in the same orders? Rn, the sorting is by largest to smallest population

```{r}
bind_cols(census1, census2)
# rows do not match because order by total pop, want it to be by state
```

```{r}
censusA <- bind_cols(
  arrange(census1, state),
  arrange(census2, state)
)

# or could have piped data in
# census1 %>% arrange(state)
# census2 %>% arrange(state)

glimpse(censusA)
```

Each dataset has a column called `state`. They both get saved but with a number. Use `.name_repair`

#### Duplicate columns?

```{r}
censusAa <- bind_cols(
  arrange(census1, state),
  arrange(census2, state)
) %>%

select(-state...4) %>%
  rename(state = state...1)

glimpse(censusAa)
```

What about all three census data?

```{r}
census_all3 <- bind_cols(
  arrange(census1, state),
  arrange(census2, state),
  arrange(census3, state)
) %>%
  select(-state...4, -state...7) %>%
  rename(state = state...1)

glimpse(census_all3) 
```

Challenge! How can I put this into a long form dataset with year as a variable?

```{r}
census_all3_longer <- census_all3 %>%
  pivot_longer(
    cols = `1970`: `2020`,
    names_to = "year",
    values_to = "population"
)

glimpse(census_all3_longer)
```


Want to add the electoral vote from the `votes` dataset to the census dataset. Want to keep everything in the census dataset (X) and add to it where there are matches in `votes` (Y).

What are the `key` variables? Both have `state` and `year` that will uniquely id rows.

### `left_join()`

```{r}
left_join(census_all3_longer, votes, by = )
```

### `right_join()`

```{r}

```

### `semi_join()`

```{r}

```

### `anti_join()`

```{r}

```

<hr>
```{r}
sessionInfo()
```